Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-06-20 papers/GPUScA-2011/foropencl.tex
Simplifications as discussed in Portland

\begin{abstract}
Emerging architectures for high performance computing are well suited
to a data-parallel programming model.  This paper presents a simple
programming methodology that allows programmers to take advantage of
these emerging systems.  We use array constructs in Fortran 90 to show how this
infrequently exploited, standardized language feature is easily
transformed to lower-level accelerator code.  These transformations
are based on a simple mapping from Fortran 90 to OpenCL.  We
demonstrate, using a shallow-water code, that an acceleration of up to 40
times can be achieved on GPU hardware with these compiler-generated
transformations.
\end{abstract}

>>>>

\begin{abstract}
Emerging GPU architectures for high performance computing are well suited to a
data-parallel programming model.  This paper presents a simple programming
methodology that allows programmers to take advantage of these emerging systems.
We use array constructs in Fortran 90 to show how this infrequently exploited,
standardized language feature is easily transformed to lower-level accelerator
code.  These transformations are based on a simple mapping from Fortran 90 to
OpenCL.  We demonstrate, using a stencil code solving the shallow-water fluid
equations, that the performance of our compiler-generated transformations is
comparable with that of hand-optimized OpenCL code.
\end{abstract}

### 2011-06-20 papers/GPUScA-2011/intro.tex
Simplifications as discussed in Portland

This paper presents a compiler-level approach for targeting a single
program to multiple, fundamentally different low-level execution
models.  This technique allows the application programmer to adopt
a single high-level programming model without sacrificing performance.
We show that features of
Fortran 90 for data-parallel programming are well suited to automatic
transformation to generate code specifically tuned for different
hardware architectures using low-level programming models such as
OpenCL and CUDA.  For algorithms that can be easily expressed in terms
of whole array, data-parallel operations, writing code in Fortran and
transforming it automatically to specific low-level implementations
removes the burden from the programmer of working with tedious, error
prone, low-level tools.

In the ideal situation, application programmers would like to adopt a
programming model in which they write their application once and use
automated tools to retarget it to many architectures.  This has proven
to be very challenging historically due to the subtle balance between
high-level expressiveness of code and the performance of the
lower-level code that is emitted by a compiler.  This ideal high-level
model that programmers work with should emphasize readability,
maintainability, and close proximity in abstraction to the problem
being solved --- in this instance, the abstraction that we care about
are mathematical formulae.  It should not be corrupted with details of
specific target architectures solely for the purpose of single-system
performance.  For certain classes of applications, specifically those
that map onto a data-parallel programming model, we show that
Fortran 90 contains language features that encourage high-level
programming abstractions without sacrificing performance during
low-level code generation.

@@

The performance that new accelerator architectures offer comes at a
cost, as processor architectures are trending toward multiple cores
with instances of integrated accelerator units (with user managed
memory) and less of a reliance on superscalar instruction level
parallelism and hardware managed memory hierarchies (such as
traditional caches).  These changes place a heavy burden on
application programmers as they are forced to adapt to the new
systems.  An especially challenging problem faced by application
programmers is not only how to program to these new architectures
(considering the massive scale of concurrency available), but also how
to design programs that are portable across the changing landscape of
computer architectures with unique memory systems and programming
models.  Can a programmer easily write one program that can run on a
conventional multicore CPU, graphics processing unit, Cell processor,
and one of many emerging many core architectures?  The fundamental
question we address is what programming model and language constructs
are best suited to span this set of new hardware designs.

@@

A common theme amongst new processors is the emphasis on data-parallel
programming.  This model is well suited to emerging architectures that
are based on either vector processing or massively parallel
collections of simple cores.  The recent CUDA and OpenCL programming
languages are are intended to support this programming model, as are
directive-based methods such as OpenMP or the Accelerator programming
model from the Portland Group~\cite{pgi10accelerator}.

@@

The problem with many of these choices is that they expose too much
detail about the machine architecture to the programmer.  This is
particularly true of CUDA and OpenCL.  In CUDA, programmers must adapt
their codes to fit the threading model used by NVIDIA GPUs, while
OpenCL requires programmers to provide specially tuned versions of
their code for different classes of machine.  In both cases, the
programmer is responsible for explicitly managing memory,
including staging of data back and forth from the host CPU and the
accelerator device memory.  While these models have been attractive as
a method for early adopters to utilize these new architectures, they
are less attractive to programmers who do not have the time or
resources to manually port their code to every new architecture and
programming model that emerges.

@@

We demonstrate that the array syntax of Fortran maps surprisingly well
onto GPUs when transformed to OpenCL kernels.  These Fortran language
features include pure and elemental functions and array constructs
like {\tt where} and {\tt cshift}.  In addition we add a few functions
that enable a program to be deployed on machines with a hierarchy of
processing elements, such as nodes employing GPU acceleration,
\emph{without requiring explicit declaration of parallelism within the
  program.}  In addition the program uses entirely standard Fortran so
it can be compiled for and executed on a single core without concurrency.
This work also is applicable to vendor-specific languages similar to
OpenCL such as the NVIDIA CUDA language.

>>>>

This paper presents a compiler-level approach for targeting a single program to
multiple and possibly fundamentally different processor architectures.  This
technique allows the application programmer to adopt a single, high-level
programming model without sacrificing performance.  We suggest that existing
data-parallel features in Fortran are well suited to applying automatic
transformations that generate code specifically tuned for different hardware
architectures using low-level programming models such as OpenCL.  For algorithms
that can be easily expressed in terms of whole array, data-parallel operations,
writing code in Fortran and transforming it automatically to specific low-level
implementations removes the burden of creating and maintaining multiple versions
of architecuture specific code.

%%%In the ideal situation, application programmers would like to adopt a programming model in which they write their application once and use automated tools to retarget it to many architectures.  This has proven to be very challenging historically due to the subtle balance between high-level expressiveness of code and the performance of the lower-level code that is emitted by a compiler.  This ideal high-level model that programmers work with should emphasize readability, maintainability, and close proximity in abstraction to the problem being solved --- in this instance, the abstraction that we care about are mathematical formulae.  It should not be corrupted with details ofspecific target architectures solely for the purpose of single-system performance.  For certain classes of applications, specifically those that map onto a data-parallel programming model, we show that Fortran 90 contains language features that encourage high-level programming abstractions without sacrificing performance during low-level code generation.

@@

Unfortunately the performance that the new accelerator architectures offer comes
at a cost.  Architectural changes are trending toward multiple heterogeneous
cores and less of a reliance on superscalar instruction level parallelism and
hardware managed memory hierarchies (such as traditional caches).  These changes
place a heavy burden on application programmers as they are forced to adapt to
these new systems.  An especially challenging problem is not only how to program
to these new architectures (considering the massive scale of concurrency
available), but also how to design programs that are portable across the
changing landscape of computer architectures with unique memory systems and
programming models.  How can a programmer easily write one program that can
perform well on both a conventional multicore CPU and a GPU (or one of the
emerging many-core architectures)?  A directive-based approach, such as OpenMP
or the Accelerator programming model from the Portland
Group~\cite{pgi10accelerator}, is one solution to this problem.

%%%The fundamental question we address is what programming model and language constructs are best suited to span this set of new hardware designs.

@@

In this paper we take a somewhat different approach.  A common theme amongst the new
processors is the emphasis on data-parallel programming.  This model is well
suited to the emerging architectures that are based on either vector processing
or massively parallel collections of simple cores.  The recent CUDA and OpenCL
programming languages are are intended to support this programming model,

@@

The problem with OpenCL and CUDA is that they expose too much detail about the
machine architecture to the programmer.  The programmer is responsible for
explicitly managing memory (including the staging of data back and forth between
the host CPU and the accelerator device) and specifically taking into account
architectural differences (such as whether the architecture contains vector
units).  While these languages have been attractive as a method for early
adopters to utilize these new architectures, they are less attractive to
programmers who do not have the time or resources to manually port their code to
every new architecture and programming model that emerges.

@@

We demonstrate that the data-parallel features of Fortran map surprisingly well
onto GPUs when transformed to OpenCL kernels.  These features include: array syntax
employing binary operators, array constructs like {\tt WHERE} and {\tt DO CONCURRENT},
and the use of pure and elemental functions.
In addition we add functions that explicitly take advantage of the
stencil geometry.

that enable a program to be deployed on machines with a hierarchy of
processing elements, such as nodes employing GPU acceleration,
\emph{without requiring explicit declaration of parallelism within the
  program.}  In addition the program uses entirely standard Fortran so
it can be compiled for and executed on a single core without concurrency.
This work also is applicable to vendor-specific languages similar to
OpenCL such as the NVIDIA CUDA language.
